---
title: "Species Distribution Modeling - Bluefun Tuna (thunnus thynnus"
author: "Johnathan Evanilla"
date: "10/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install and load necessary packages

```{r warning=FALSE, message=FALSE}
# Install packages
installed <- rownames(installed.packages())
if (!("robis" %in% installed)){
   if (!("remotes" %in% installed))install.packages("remotes")
   remotes::install_github("iobis/robis")
}
if (!("ohwobpg" %in% installed)){
  remotes::install_github("BigelowLab/ohwobpg", quiet = TRUE)
}

# Load libraries
library(robis, quietly = TRUE)
library(ohwobpg, quietly = TRUE)
library(dplyr, quietly = TRUE)
library(biomod2, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(raster, quietly = TRUE)
library(viridis, quietly = TRUE)
```

## Fetching Thunnus thynnus occurences

```{r fetch_occurrences}
T.thynnus <- robis::occurrence("Thunnus thynnus")
```

## Plot occurrences

Plotting the Thunnus thynnus occurrences, again, I've just used the example and put in Thunnus thynnus

```{r}
# Load world map data 
worldmap <- ggplot2::map_data("world")

# Plot occurrences on world map
ggplot2::ggplot(data = T.thynnus, mapping = aes(x = decimalLongitude, y = decimalLatitude)) +
  # Add occurrence data
  geom_point() +
  # Add map data
  geom_polygon(data = worldmap, aes(long, lat, group = group), fill = NA, colour = "gray43") +
        coord_quickmap(xlim = c(round(min(T.thynnus$decimalLongitude)), 
                                round(max(T.thynnus$decimalLongitude))), 
                       ylim = c(round(min(T.thynnus$decimalLatitude)), 
                                round(max(T.thynnus$decimalLatitude))),
                       expand = TRUE) +
  # Clean up theme
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```
## Load OBPG environmental data

Next, we will load the environmental data layers we will be using in our model.  This is done using the ohwobpg package that was loaded earlier.  This example focuses on the Gulf of Maine, to follow the previous example, using sea surface temperature (sst) and chlorophyll A (chlor_a).

The code below needs a little more explanation. I think it's pulling data from summer 2018.

```{r}
# Load database
path <- system.file("gom", package = "ohwobpg")
db <- read_database(path)

# Load sst ----
sst_db <- db %>% 
  dplyr::filter(param == "sst" & 
                per == "MO" &
                dplyr::between(date, as.Date("2018-05-15"), as.Date("2018-09-26")))

# Create raster stack
sst <- sst_db %>%                # start with the subset database
  as_filename(path = path) %>%   # build filenames and append to the path
  raster::stack()                # read them into a stack of images
 
# Name layers
names(sst) <- format(sst_db$date, "%b")

# Load chlor_a ----
chlor_a_db <- db %>% 
  dplyr::filter(param == "chlor_a" & 
                per == "MO" &
                dplyr::between(date, as.Date("2018-05-15"), as.Date("2018-09-26")))

# Create raster stack
chlor_a <- chlor_a_db %>%                # start with the subset database
  as_filename(path = path) %>%   # build filenames and append to the path
  raster::stack()                # read them into a stack of images
 
# Name layers
names(chlor_a) <- format(chlor_a_db$date, "%b")

# Remove lakes from SST ----
sst[is.na(chlor_a)] <- NA

```

## Subset T. thynnus data

Next, we will subset the M. mola data so it's in the same bounding box as our environmental data.  We will also subset to only include occurrence records from August, matching the environmental data above and the example. Will tinker with this later.

```{r}
T.thynnus.subset <- T.thynnus %>% 
  # Shorten names
  dplyr::rename(lon = decimalLongitude,
                lat = decimalLatitude) %>%
  # Filter to bounding box and select month of August
  dplyr::filter(lon >= raster::xmin(chlor_a) & lon <= raster::xmax(chlor_a) &
                lat >= raster::ymin(chlor_a) & lat <= raster::ymax(chlor_a) &
                month == "08") %>%
  # Select relevant columns
  dplyr::select(lon, lat)
                
```

## Add background points

We need to add a column for binary presence/absence in order to model the species' distribution with biomod2. Because the dataset only includes presences, we first need to select random backgroun points.  Next, we will add a column containing a 1 to our occurrence dataset and a column containing a zero to the background points dataset.  Then, the two datasets are combined into one presence/absence dataframe.

```{r}
# Add presence/absence code
# Add unique latitude and longitude code
T.thynnus.biomod <- T.thynnus.subset %>%
  dplyr::mutate(pa = 1, code = paste0(lon, lat))

# Add background points
# Select points using unique latitude and longitude code
background_locs <- as.data.frame(sst$Aug, xy = TRUE) %>% 
  dplyr::rename(lon = x, lat = y) %>%
  sample_n(500) %>%
  dplyr::mutate(pa = 0, code = paste0(lon, lat)) %>%
  dplyr::filter(!(code %in% T.thynnus.biomod$code) & !is.na(Aug)) %>%
  dplyr::select(lon, lat, pa, code)

# Combine occurrence data and background points
T.thynnus.biomod <- rbind(T.thynnus.biomod, background_locs)

```


## Plot occurrences with background points

Next, we will plot the occurrences again to ensure they are within the bounding box and do not overlap with the background points.  To determine this, we will plot the points on top of the SST layer.

```{r}
# Plot occurrences and background points on top of sst data
ggplot2::ggplot() + 
  geom_tile(data = as.data.frame(sst$Aug, xy = TRUE), mapping = aes(x = x, y = y, fill = Aug)) +
  scale_fill_viridis(na.value = "white") +
  # Add occurrence data
  geom_point(data = T.thynnus.subset, mapping = aes(x = lon, y = lat), color = "blue") +
  # Add background data
  geom_point(data = background_locs, mapping = aes(x = lon, y = lat), color = "red") +
  # Add map data
  geom_polygon(data = worldmap, aes(long, lat, group = group), fill = NA, colour = "gray43") +
  coord_quickmap(xlim = c(xmin(sst), 
                          xmax(sst)), 
                 ylim = c(ymin(sst), 
                          ymax(sst)),
                 expand = TRUE) +
  # Clean up theme
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

In the E.glacialis example, there was a point on land that needed to be removed. I don't see any such cases here, so I'll comment out the next block.

```{r}
#E.glacialis.biomod <- E.glacialis.biomod %>%
#  dplyr::filter(!(pa == 1 & lat > 45.5))
```

Now the data is ready for use in building our species distribution model using a package called [biomod2](https://cran.r-project.org/web/packages/biomod2/biomod2.pdf).

## Format the data for biomod2

First, we need to format the data for use in biomod2 with the species name, the presence or absence column, and the spatial coordinates as separate variables.

```{r}
# Name of studied species
species_name <- "T.thynnus"
# Presence data for the species
species_occurrence <- as.numeric(unlist(T.thynnus.biomod[, "pa"]))
# XY coordinates of species data
species_xy <- T.thynnus.biomod[, c("lon","lat")]
# Environmental covariates
env_covars <- raster::stack(sst$Aug, chlor_a$Aug)
```

Next, convert the data into a biomod2 formatted data object.

```{r}
# Format using biomod2 function
formatted_data <- biomod2::BIOMOD_FormatingData(resp.var = species_occurrence,
                                                expl.var = env_covars,
                                                resp.xy = species_xy,
                                                resp.name = species_name)

# Check data
formatted_data
```

## Define modeling options

Next, define the modeling options using the biomod2 built-in function.  This is where individual modeling algorithm parameters are defined.  For the purpose of this tutorial, we are using the default values.

```{r}
# Initialize modeling options
# This is where model parameters are defined
# Leaving as default for this tutorial
modeling_options <- BIOMOD_ModelingOptions()
```

## Build the model

Then, build the model. While biomod2 has the capability to model using 10 species distribution modeling algorithms, for the purpose of this tutorial, we will only build a generalized additive model, or a GAM.

```{r}
# Build model
model <- BIOMOD_Modeling(data = formatted_data,
                         models = c("GAM"),
                         models.options = modeling_options,
                         NbRunEval = 3,
                         # Implement a 70%-30% train/evaluation data split
                         DataSplit = 70,
                         models.eval.meth = c("ROC"),
                         do.full.models = FALSE,
                         modeling.id = paste0(species_name,"Modeling"))

# Check summary
model
```

## Assess the model

To see how well the model performed, we can assess the evaluations.  For this model, we used the area under the receiver operator characteristic curve, or ROC, as the evaluation metric.  This metric is on a scale of 0 to 1, with 0.5 meaning the model performed as well as a random model, and anything above 0.5 indicating the model performed better than a random model.

```{r}
# Access evaluations
evals <- get_evaluations(model)

# Assess the area under the receiver operator characteristic curve (ROC)
evals["ROC","Testing.data","GAM",,]
```

From the evaluations, it looks like our models performed relatively well.

## Project the model

Finally, we will project the model back onto the original environmental data layers.  This just means we will be using our model to make a guess, or projection, about the probability of M.mola being present in any given grid cell within our domain.

```{r}
# Project model
proj <- BIOMOD_Projection(modeling.output = model,
                          new.env = env_covars, 
                          proj.name = 'current', 
                          selected.models = 'all', 
                          binary.meth = 'ROC', 
                          compress = 'xz', 
                          clamping.mask = F, 
                          output.format = '.grd')

# See where projection files are stored
list.files("T.thynnus/proj_current/")

# Load projection as a dataframe
proj_df <- raster::raster(file.path("T.thynnus/proj_current/proj_current_T.thynnus.grd")) %>%
  # Biomod2 computes probability 1000x higher than it actually is to save memory, so dividing
  # by 1000 will convert the probability back to a 0-1 scale
  `/`(1000) %>%
  as.data.frame(xy = TRUE) %>%
  dplyr::rename(prob = T.thynnus_AllData_RUN1_GAM)

#Plot map, probabilities, and sightings data
ggplot() + 
  # Add probability data
  geom_tile(data = proj_df, aes(x, y, fill = prob)) +
  # Add probabilitiy color gradient and label
  scale_fill_gradientn(colours = inferno(500), limits = c(0,1), name = '', na.value = "white") +
  # Add sightings data
  geom_point(data = T.thynnus.biomod %>% dplyr::filter(pa == 1), mapping = aes(x = lon, y = lat), color = "white", fill = "black", pch = 21, cex = 1, stroke = 0.5) +
  labs(x = "", 
       y = "", 
       title = "August 2018 Bluefin Tuna Distribution") +
  # Add world map data
  geom_polygon(data = worldmap, aes(long, lat, group = group), fill = NA, colour = "gray43") +
  coord_quickmap(xlim = c(round(min(proj_df$x)), round(max(proj_df$x))), 
                 ylim = c(round(min(proj_df$y)), round(max(proj_df$y))),
                 expand = TRUE) +
  # Remove grid lines
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

The regions in yellow represent the parts of the domain where the model projects high probability of an M.mola occurrence.  The regions in black indicate that the model projected a low probability of occurrence. 

This model looks like it misses the typical late summer distribution of Bluefin in coastal waters within the Gulf of Maine, specifically off of Cape Cod.

